Neural Connectome of the Ctenophore Statocyst

### 2025-06-02 manuscript/Manuscript.qmd
ms edits, add refs

Ctenophores, or comb jellies, are gelatinous marine animals that actively swim by beating rows of fused cilia known as comb plates. These coordinated ciliary movements not only generate propulsion but also allow precise control of body orientation and directional changes within the water column. Despite lacking a centralized nervous system, ctenophores exhibit sophisticated and dynamic behavioural patterns and respond to a broad range of external stimuli, suggesting the presence of an evolutionarily unique and functionally integrated sensorimotor system. Several components of the ctenophore nervous system have been described, including a subepithelial nerve net (SNN), mesogleal neurons, tentacular nerves, and elements of nervous system in the aboral organ [@Hertwig1880; @Hernandez_Nicaise_1984; @Jager2011]. These neural structures are characterized by unique features such as a syncytial architecture [@Burkhardt2023] and an extensive repertoire of lineage-specific neuropeptides not found in other metazoans [@Sachkova2021; @Hayakawa2022]. Notably, early genomic and single-cell transcriptomic studies failed to identify conserved neuronal markers in ctenophores [@Moroz2014; @Sebe-Pedros2018]. It was only through the use of pro-neuropeptide-based markers that distinct neural cell types were molecularly identified [@Sachkova2021; @Hayakawa2022]. In parallel, phylogenomic and chromosomal synteny analyses have suggested that ctenophores may represent the sister group to all other animals [@Ryan2013; @Whelan2015a; @Li2021; @Schultz2023], raising the possibility that their nervous system evolved independently from those of other metazoans [@moroz2015]. This hypothesis positions ctenophores as a key lineage for re-evaluating the origins and diversity of neural systems. However, without a detailed understanding of how ctenophore neurons are spatially organized and interconnected, it remains unclear how their nervous system controls behaviour. Clarifying this relationship between neural architecture and behavioural control is essential to uncovering the logic of ctenophore neural function.

To address the lack of structural, circuit-level understanding of the ctenophore nervous system, we focused on a unique sensory structure—the aboral organ. This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for the beating of comb plate cilia and thereby control body orientation [@HORRIDGE1965; @Tamm1980]. Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez_Nicaise1968; @Jager2011]. A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency. Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system, not a fixed one like in vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @tamm2014b]. Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997]. In particular, deflection-induced excitation can be abolished by calcium channel inhibitors or calcium-free conditions [@Lowe1997], while high extracellular K⁺ depolarizes and activates cilia in the absence of mechanical stimuli [@Lowe1997]. Local calcium application further suggests that Ca²⁺ influx at the ciliary base may be required for the excitatory response [@Lowe1997]. These findings support the hypothesis that neural modulation plays a critical role in shaping balancer-cell output [@Lowe1997].

To investigate how balancer cilia are modulated within the aboral organ, we combined volume electron microscopy and high-speed imaging in the ctenophore *Mnemiopsis leidyi*. Through connectomic reconstruction, we characterized the synaptic architecture of the aboral nerve net, revealing synaptic inputs to the balancer cell. By functional imaging, we uncovered correlated activity patterns in balancer cilia across the four quandrants of the apical organ. Our results suggest that the aboral nerve net functions in coordinating the activity of balancer cilia, rather than fulfilling a feedforward sensory-motor function.

@@

For vEM analysis, we used a five-day-old cydippid larva of *Mnemiopsis leidyi*. To obtain high-quality ultrastructural preservation, we used high pressure freezing followed by freeze substitution and resin embedding (Epon). The specimen was sectioned from the aboral tip and we collected approximately 1,000 ultra-thin (50 nm) serial sections as ribbons. The sections were imaged in a Zeiss Gemini 500 scanning electron microscope at 2.0 nm/pixel (xy). We imaged only the region containing the aboral organ. Subsequently, images from 620 layers were stitched and aligned in TrakEM2 [@cardona2012]. The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009], resulting in a reconstruction of 909 cells that had intact nuclei and cell bodies. Most of the cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ.

In cells containing cilia, we also traced cilia along their length and annotated all basal bodies. For neuronal skeletonization, nodes were placed to interconnect the profiles of the same neuron’s processes across layers, extending the skeleton until all branches were fully traced. All nodes relevant to synaptic sites were tagged, and skeletons were named and assigned multi-level annotations. As described later, some neurons formed loop-like structures (anastomosed neurons), wherein separated branches often rejoined either the main branch or smaller branches [@Burkhardt2023]. In such cases, branch nodes were placed near the closest existing node and annotated accordingly (CATMAID only supports skeleton trees). The entire skeletonized volume was composed of 134,591 nodes. 88 fragments could not be attached to somata-associated skeletons. Most of these fragments represent short skeletal branches that could not be traced beyond gaps or low-quality layers.

Next, we divided the entire aboral organ into four quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1G) [@Martindale1999]. The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores [@Martindale1999]. This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development [@Martindale1999]. We assigned each of the 909 skeletonised cells to one of the four quadrants. <!-- check number of cells, 927 in figure -->

>>>>

Ctenophores, or comb jellies, are gelatinous marine animals that actively swim by beating rows of fused cilia known as comb plates. These coordinated ciliary movements not only generate propulsion but also allow precise control of body orientation and directional changes within the water column. Despite lacking a centralized nervous system, ctenophores exhibit sophisticated and dynamic behavioural patterns and respond to a broad range of external stimuli, suggesting the presence of an evolutionarily unique and functionally integrated sensorimotor system. Several components of the ctenophore nervous system have been described, including a subepithelial nerve net (SNN), mesogleal neurons, tentacular nerves, and elements of nervous system in the aboral organ [@Hertwig1880; @Hernandez_Nicaise_1984; @Jager2011]. These neural structures are characterized by unique features such as a syncytial architecture [@Burkhardt2023] and an extensive repertoire of lineage-specific neuropeptides not found in other metazoans [@Sachkova2021; @Hayakawa2022]. Notably, early genomic and single-cell transcriptomic studies failed to identify conserved neuronal markers in ctenophores [@Moroz2014; @Sebe-Pedros2018]. It was only through the use of pro-neuropeptide-based markers that distinct neural cell types were molecularly identified [@Sachkova2021; @Hayakawa2022]. In parallel, phylogenomic and chromosomal synteny analyses have suggested that ctenophores may represent the sister group to all other animals [@Ryan2013; @Whelan2015a; @Li2021; @Schultz2023], raising the possibility that their nervous system evolved independently from those of other metazoans [@moroz2015]. This hypothesis positions ctenophores as a key lineage for re-evaluating the origins and diversity of neural systems [@Burkhardt_2021]. However, without a detailed understanding of how ctenophore neurons are spatially organized and interconnected, it remains unclear how their nervous system controls behaviour. Clarifying this relationship between neural architecture and behavioural control is essential to uncovering the logic of ctenophore neural function.

To address the lack of structural, circuit-level understanding of the ctenophore nervous system, we focused on a unique sensory structure—the aboral organ. This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for the beating of comb-plate cilia and thereby control body orientation [@HORRIDGE1965; @Tamm1980]. Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez_Nicaise1968; @Jager2011]. A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency. Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system, not a fixed one like in vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @tamm2014b]. Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997]. In particular, deflection-induced excitation can be abolished by calcium channel inhibitors or calcium-free conditions [@Lowe1997], while high extracellular K⁺ depolarizes and activates cilia in the absence of mechanical stimuli [@Lowe1997]. Local calcium application further suggests that Ca²⁺ influx at the ciliary base may be required for the excitatory response [@Lowe1997]. These findings support the hypothesis that ionic currents and potential neural inputs play critical roles in shaping balancer-cell output [@Lowe1997].

To investigate how balancer cilia are modulated within the aboral organ, we combined volume electron microscopy and high-speed imaging in the ctenophore *Mnemiopsis leidyi*. Through connectomic reconstruction, we characterized the synaptic architecture of the aboral nerve net, revealing synaptic inputs to the balancer cells. By functional imaging, we uncovered correlated activity patterns in balancer cilia across the four quandrants of the apical organ. Our results suggest that the aboral nerve net functions in coordinating the activity of balancer cilia, rather than fulfilling a feedforward sensory-motor function.

@@

For vEM analysis, we used a five-day-old cydippid larva of *Mnemiopsis leidyi*. To obtain high-quality ultrastructural preservation, we used high pressure freezing followed by freeze substitution and resin embedding (Epon). The specimen was sectioned from the aboral tip and we collected approximately 1,000 ultra-thin (50 nm) serial sections as ribbons. The sections were imaged in a Zeiss Gemini 500 scanning electron microscope at 2.0 nm/pixel (xy). We imaged only the region containing the aboral organ. Subsequently, images from 620 layers were stitched and aligned in TrakEM2 [@cardona2012]. The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009; @Schneider-Mizell2016], resulting in a reconstruction of 972 cells that had intact nuclei and cell bodies, 899 of which formed the aboral organ (AO). Most of the 899 AO cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ.

In cells containing cilia, we also traced cilia along their length and annotated all basal bodies. For neuronal skeletonization, nodes were placed to interconnect the profiles of the same neuron’s processes across layers, extending the skeleton until all branches were fully traced. All nodes relevant to synaptic sites were tagged, and skeletons were named and assigned multi-level annotations. As described later, some neurons formed loop-like structures (anastomosed neurons), wherein separated branches often rejoined either the main branch or smaller branches [@Burkhardt2023]. In such cases, branch nodes were placed near the closest existing node and annotated accordingly (CATMAID only supports skeleton trees). The entire skeletonized volume was composed of 134,591 nodes. 73 fragments could not be attached to somata-associated skeletons. Most of these fragments represent short skeletal branches that could not be traced beyond gaps or low-quality layers.

Next, we divided the entire aboral organ into four quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1G) [@Martindale1999]. The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores [@Martindale1999]. This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development [@Martindale1999]. We assigned each of the 899 skeletonised cells to one of the four quadrants (Figure 1G). <!-- check number of cells, 927 in figure -->
